Bucket-Sorted Independent Sets for Algebraic Multigrid

### 2008-02-19 bsis_paper/paper.tex
changed some wordings and comment style
git-svn-id: file:///home/rblake/svn/pamg/trunk@42 c7f3fe13-8c30-0410-9127-9a9c42fe009c

\usepackage{amssymb,listings}

@@

\usepackage{listings}
\lstset{mathescape,
        escapechar=\%,
        numbers=left, 
        numberstyle=\tiny, 
        stepnumber=1, 
        numbersep=5pt,
        frame=tb,
        xleftmargin=5mm,
        framexleftmargin=5mm,
        language=Matlab,
        belowcaptionskip=5pt}
\renewcommand{\lstlistingname}{Algorithm}
\usepackage{caption}
\captionsetup[lstlisting]{justification=raggedright}

@@

\begin{abstract}
Independent set-based coarse-grid selection algorithms are defined by
policies for weight initialization, independent set selection, and
weight update. In this paper, we develop theory demonstrating that
algorithms employing the same policies produce identical coarse grids,
regardless of the implementation. This result on coarse grid
invariance is applied to produce a new coarse-grid selection algorithm
that is substantially more efficient than an existing algorithm using
the same policies. Experimental results comparing two versions of the
new algorithm and the older algorithm are presented, followed by a
discussion on possible approaches to effectively parallelize the new
method.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
The algebraic multigrid (AMG) method~\cite{Ruge1987,Brandt1982} is an
efficient numerical algorithm for finding the solution to numerous
problems, including linear systems of the form $Ax = b$ arising from
the discretization of partial differential equations on structured and
unstructured meshes. In many cases, the computational complexity of
AMG is $O(n)$, where $n$ is the number of unknowns in the linear
system. The linear cost scaling property of AMG makes the method
attractive, especially for large problems.

@@

In this paper, we focus attention on coarse-grid selection algorithms
in the AMG setup phase. The independent set-based class of coarsening
algorithms is studied in detail. New theory is developed to
demonstrate invariance in the result of independent set-based coarse
grid selection, given restrictions on the selection
policies. Algorithms that fall within the assumptions of the theory
are guaranteed to produce the same coarse grids, regardless of how the
policies are implemented. A new coarse-grid selection algorithm,
designed to select coarse grids more efficiently than the CLJP-c
algorithm, is presented to demonstrate the theoretical results. A
particularly attractive result of the theory is that computational
efficiency concerns become decoupled from the output of coarse-grid
selection.

The remainder of this paper is organized as follows. In
Section~\ref{sec:cgs}, we discuss contemporary coarse-grid selection
algorithms. The components of independent set-based coarsening
algorithms are presented. Section~\ref{sec:invariance} develops theory
based on general independent set-based coarsening and the invariance
of coarse grids selected when the same selection policies are
used. Section~\ref{sec:bsis} demonstrates the results from
Section~\ref{sec:invariance} through the introduction of a new
coarsening algorithm called Bucket-Sorted Independent Sets
(BSIS). BSIS produces the same coarse grids as the CLJP-c coarsening
algorithm, but is designed to do so in a much more efficient
manner. Experimental results are reported to demonstrate the
performance gains in BSIS, and discussion is included to propose a
method for the parallelization of BSIS. Conclusions are presented in
Section~\ref{sec:conclusions}.

@@

Given a set of fine-level degrees of freedom, $\Omega_k$, nodal
coarse-grid selection constructs a set of coarse-level degrees of
freedom $\Omega_{k+1} \subset \Omega_k$. Several approaches have been
developed to select the coarse grid, with independent set-based
approaches being the most numerous. Independent set-based algorithms
descend from the classical AMG coarsening algorithm, Ruge-St\"uben
(RS) coarsening~\cite{Ruge1987}, also called classical
coarsening. Independent set-based methods are our focus and are
discussed in more detail in the next section.

@@

Each vertex in the graph is assigned a weight based on its merit for
becoming a $C$-point. The largest weights are conferred to vertices
best suited to be $C$-points, where the quality of potential
$C$-points is quantified by strength of connection measures.

@@

The weight update routines in independent set-based coarsening
algorithms (encapsulated in Line~8 of
Algorithm~\ref{alg:general-indep-set}) are varied but take one of two
forms: routines that decrease the values of weights and routines that
increase the value of weights. In all cases, a vertex with a large
weight has high fitness and is more desirable as a $C$-point. As the
pseudocode is formatted in Algorithm~\ref{alg:general-indep-set},
assignment of vertices to the $F$-point set occurs during the weight
update routine.

@@

Given a symmetric, but otherwise arbitrary, set of selection
neighborhoods, the set of vertices potentially able to affect the
vertex weight of $i$ or $j \in \mathcal{N}_i^s$ is the \emph{extended
selection neighborhood}.
\begin{defn}
The \emph{extended selection neighborhood} of $i$, denoted
$\mathcal{N}_i^{2s}$, is the union of the selection neighborhood of
$i$ with the selection neighborhoods of the vertices in
$\mathcal{N}_i^s$, excluding $i$. That is, $\mathcal{N}_i^{2s} =
\left[\left(\bigcup_{j \in \mathcal{N}_i^s}\mathcal{N}_j^s\right)\cup
\mathcal{N}_i^s\right] \setminus \{i\}$.
\end{defn}

When a vertex $i$ satisfies the generalized selection
condition~(\ref{eqn:general-selection}), no other valid $C$-point
assignments are able to affect $w_i$. This is formalized below.
\begin{lem}
\label{5:lemma:C-lock}
If (\ref{eqn:general-selection}) is satisfied for vertex $i$, then $i$
must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become a
$C$-point, regardless of other $C$-point selections and corresponding
weight updates made in the graph.
\end{lem}
\begin{proof}
The satisfaction of (\ref{eqn:general-selection}) means $w_i > w_j$
for all $j \in \mathcal{N}_i^s$. For $i$ to not become a $C$-point,
its weight must become smaller than some $j \in \mathcal{N}_i^s$. The
weight of $i$, however, is not decremented unless some $j \in
\mathcal{N}_i^s$ satisfies~(\ref{eqn:general-selection}) and becomes a
$C$-point, which is impossible until $i$ is assigned to the $C$-point
set.
\end{proof}

@@

Lemma~\ref{5:lemma:D0_invariant} states that any coarse-grid selection
method using the general selection condition invariably selects $D_0$
vertices as $C$-points. This result is used to prove, given the
conditioned developed earlier in this section, the result of
coarse-grid selection is insensitive to the implementation decisions
made in specific algorithms.

@@

Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{alg:setup}.
\begin{lstlisting}[caption={CLJP Weight Update for CSR Matrix},label=alg:setup]
%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$%
for $d \in D$
  for $i \in S_d$
    $w_i \leftarrow w_i - 1$
    $S_d = S_d \setminus \{i\}$ //removing edge from graph
  end
end
for $i \notin (C \cup F)$
  for $k$ originally in $S_i$
    if $k \in D$
      mark $k$ NEW-C-POINT
    end
  end
  for $j \in S_i$
    if $j \notin D$
      for $k \in S_j$
        if $k$ is marked NEW-C-POINT      // $i$ and $j$ mutually influenced by $k$
          $w_j \leftarrow w_j - 1$
          $S_i = S_i \setminus \{j\}$     // remove edge from $j$ to $i$
        end
      end
    end
  end
  for $k$ originally in $S_i$
    if $k \in D$
      unmark $k$
      $S_i = S_i \setminus \{k\}$         // remove edge from $k$ to $i$, if present
    end
  end
end
\end{lstlisting}
%\begin{algorithm}[h]
%\caption{CLJP Weight Update for CSR Matrix}
%\label{alg:setup}
%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
%\begin{algorithmic}[1]
%\FORALL{$d \in D$}
%\FORALL{$i \in S_d$}
%\STATE $w_i \leftarrow w_i - 1$ \COMMENT{see
%  Figure~\ref{3:fig:CLJP-update-heuristics} left}
%\STATE $S_d = S_d \setminus \{i\}$ \COMMENT{removing edge from graph}
%\ENDFOR
%\ENDFOR
%\FORALL{$i \notin (C \cup F)$}
%\FORALL{$k$ originally in $S_i$}
%\IF{$k \in D$}
%\STATE mark $k$ NEW-C-POINT
%\ENDIF
%\ENDFOR
%\FORALL{$j \in S_i$}
%\IF{$j \notin D$}
%\FORALL{$k \in S_j$}
%\IF[$i$ and $j$ mutually influenced by $k$]{$k$ is marked NEW-C-POINT}
%\STATE $w_j \leftarrow w_j - 1$ \COMMENT{see
%  Figure~\ref{3:fig:CLJP-update-heuristics} right}
%\STATE $S_i = S_i \setminus \{j\}$ \COMMENT{remove edge from $j$ to $i$}
%\ENDIF
%\ENDFOR
%\ENDIF
%\ENDFOR
%\FORALL{$k$ originally in $S_i$}
%\IF{$k \in D$}
%\STATE unmark $k$
%\STATE $S_i = S_i \setminus \{k\}$ \COMMENT{remove edge from $k$ to $i$,
%  if present}
%\ENDIF
%\ENDFOR
%\ENDFOR
%\end{algorithmic}
%\}
%\end{algorithm}
The level of complication in this update routine is due to the CSR
format and the need to find vertices strongly influenced by new
$C$-points. When a new $C$-point $k$ is selected, the first type of
weight update (Lines 2--7) is trivial since determining the vertices
in $S_k$ is inexpensive. The second type of update (Lines 8--30) is
more expensive since the vertices influenced by $k$ are difficult to
determine in a CSR format. The update requires a search of many
vertices $i$ and all of their strong influencing neighbors $j$. The
routine then searches strongly influencing $j$ to determine if any $k
\in D$ strongly influences both $i$ and $j$. The cost increases
dramatically as the density of $S$ increases. Large operator
complexities have a disproportionately large impact on coarse-grid
selection run time. In the next section, a modified update routine to
compliment the new search technique is introduced.

>>>>

\usepackage{amssymb,listings,color}
\definecolor{mygray}{gray}{0.6}

@@

\usepackage{listings}
\lstset{mathescape,
        escapechar=\%,
        numbers=left, 
        numberstyle=\tiny, 
        stepnumber=1, 
        numbersep=5pt,
        frame=tb,
        xleftmargin=5mm,
        framexleftmargin=5mm,
        language=Matlab,
        morecomment=[n][\itshape\color{mygray}]{/*}{*/},
        belowcaptionskip=5pt}
\renewcommand{\lstlistingname}{Algorithm}
\usepackage{caption}
\captionsetup[lstlisting]{justification=raggedright}

@@

\begin{abstract} Independent set-based coarse-grid selection algorithms for
algebraic multigrid are defined by their policies for weight initialization,
independent set selection, and weight update. In this paper, we develop theory
demonstrating that algorithms employing the same policies produce identical
coarse grids, regardless of the implementation. The coarse grid invariance
motivates a new coarse-grid selection algorithm that is more efficient than an
existing algorithm using the same policies. Experimental results highlighting
the efficiency of two versions of the new algorithm are presented, followed by a
discussion on possible approaches to effectively parallelize the new method.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
The algebraic multigrid (AMG) method~\cite{Ruge1987,Brandt1982} is an
efficient numerical algorithm for finding the solution to
linear systems of the form $Ax = b$.  Often, these algebraic systems arise from
the discretization of partial differential equations on structured and
unstructured meshes. In many cases, the computational complexity of
AMG is $O(n)$, where $n$ is the number of unknowns in the linear
system. The linear cost property of AMG makes the method particularly
attractive for large scale problems.

@@

In this paper, we focus attention on coarse-grid selection algorithms
in the AMG setup phase. The independent set-based class of coarsening
algorithms is studied in detail. New theory is developed to
demonstrate invariance in the result of independent set-based coarse
grid selection, given restrictions on the selection
policies. Algorithms that fall within the assumptions of the theory
are guaranteed to produce the same coarse grids, regardless of how the
policies are implemented. A new coarse-grid selection algorithm,
designed to select coarse grids more efficiently than the CLJP-c
algorithm, is presented to demonstrate the theoretical results. A
particularly attractive result of the theory is that output from coarse-grid
selection may be considered independent of the computational efficiency.

The remainder of this paper is organized as follows. In
Section~\ref{sec:cgs}, we discuss contemporary coarse-grid selection
algorithms. The components of independent set-based coarsening
algorithms are presented. Section~\ref{sec:invariance} develops theory
based on general independent set-based coarsening and the invariance
of coarse grids selected when the same selection policies are
used. Section~\ref{sec:bsis} demonstrates the results from
Section~\ref{sec:invariance} through the introduction of a new
coarsening algorithm called Bucket-Sorted Independent Sets
(BSIS). BSIS produces the same coarse grids as the CLJP-c coarsening
algorithm, but is designed to do so in a more efficient
manner. Experimental results are reported to demonstrate the
performance gains in BSIS, and discussion is included to propose a
method for the parallelization of BSIS. Conclusions are presented in
Section~\ref{sec:conclusions}.

@@

Given a set of fine-level degrees of freedom, $\Omega_k$, nodal
coarse-grid selection constructs a set of coarse-level degrees of
freedom $\Omega_{k+1} \subset \Omega_k$. Several approaches have been
developed to select the coarse grid, with independent set-based
approaches being the most numerous. Independent set-based algorithms
descend from the classical AMG coarsening algorithm, Ruge-St\"uben
(RS) coarsening~\cite{Ruge1987}, also called classical
coarsening. Independent set-based methods are the focus of this paper and are
discussed in more detail in the next section.

@@

Each vertex in the graph is assigned a weight based on its merit in
becoming a $C$-point. The largest weights are associated with vertices
best suited to be $C$-points, where the quality of potential
$C$-points is quantified by strength of connection measures.

@@

The weight update routines in independent set-based coarsening
algorithms (cf. Line~8 of
Algorithm~\ref{alg:general-indep-set}) vary, but take one of two
forms: routines that decrease the values of weights and routines that
increase the value of weights. In all cases, a vertex with a large
weight has high fitness and is more desirable as a $C$-point. The
pseudocode as written in Algorithm~\ref{alg:general-indep-set} performs
assignment of vertices to the $F$-point set during the weight
update routine.

@@

Given a symmetric, but otherwise arbitrary, set of selection
neighborhoods, the set of vertices affecting the
vertex weight of $i$ or $j \in \mathcal{N}_i^s$ is the \emph{extended
selection neighborhood}.
\begin{defn}
The \emph{extended selection neighborhood} of $i$, denoted
$\mathcal{N}_i^{2s}$, is the union of the selection neighborhood of
$i$ with the selection neighborhoods of the vertices in
$\mathcal{N}_i^s$, excluding $i$. That is, $\mathcal{N}_i^{2s} =
\left[\left(\bigcup_{j \in \mathcal{N}_i^s}\mathcal{N}_j^s\right)\cup
\mathcal{N}_i^s\right] \setminus \{i\}$.
\end{defn}

When a vertex $i$ satisfies the generalized selection
condition~(\ref{eqn:general-selection}), no other valid $C$-point
assignments are able to affect $w_i$. This is formalized below.
\begin{lem}
\label{5:lemma:C-lock}
If (\ref{eqn:general-selection}) is satisfied for vertex $i$, then $i$
is the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become a
$C$-point, regardless of other $C$-point selections and corresponding
weight updates made in the graph.
\end{lem}
\begin{proof}
The satisfaction of (\ref{eqn:general-selection}) means $w_i > w_j$
for all $j \in \mathcal{N}_i^s$. For $i$ to not become a $C$-point,
its weight must become smaller than some $j \in \mathcal{N}_i^s$. The
weight of $i$, however, is not decremented unless some $j \in
\mathcal{N}_i^s$ satisfies~(\ref{eqn:general-selection}) and becomes a
$C$-point, which is impossible until $i$ is assigned to the $C$-point
set.
\end{proof}

@@

Lemma~\ref{5:lemma:D0_invariant} states that any coarse-grid selection
method using the general selection condition invariably selects $D_0$
vertices as $C$-points. This result is used to prove, given the
conditioned developed earlier in this section, that of
coarse-grid selection is insensitive to the implementation decisions
made in specific algorithms.

@@

Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{alg:setup}.
\begin{lstlisting}[caption={CLJP Weight Update for CSR Matrix},label=alg:setup]
%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$%
for $d \in D$
  for $i \in S_d$
    $w_i \leftarrow w_i - 1$
    $S_d = S_d \setminus \{i\}$ /*removing edge from graph*/
  end
end
for $i \notin (C \cup F)$
  for $k$ originally in $S_i$
    if $k \in D$
      mark $k$ NEW-C-POINT
    end
  end
  for $j \in S_i$
    if $j \notin D$
      for $k \in S_j$
        /* $i$ and $j$ mutually influenced by $k$ */
        if $k$ is marked NEW-C-POINT      
          $w_j \leftarrow w_j - 1$
          $S_i = S_i \setminus \{j\}$     /*remove edge from $j$ to $i$*/
        end
      end
    end
  end
  for $k$ originally in $S_i$
    if $k \in D$
      unmark $k$
      $S_i = S_i \setminus \{k\}$         /*remove edge from $k$ to $i$, if present*/
    end
  end
end
\end{lstlisting}
%\begin{algorithm}[h]
%\caption{CLJP Weight Update for CSR Matrix}
%\label{alg:setup}
%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
%\begin{algorithmic}[1]
%\FORALL{$d \in D$}
%\FORALL{$i \in S_d$}
%\STATE $w_i \leftarrow w_i - 1$ \COMMENT{see
%  Figure~\ref{3:fig:CLJP-update-heuristics} left}
%\STATE $S_d = S_d \setminus \{i\}$ \COMMENT{removing edge from graph}
%\ENDFOR
%\ENDFOR
%\FORALL{$i \notin (C \cup F)$}
%\FORALL{$k$ originally in $S_i$}
%\IF{$k \in D$}
%\STATE mark $k$ NEW-C-POINT
%\ENDIF
%\ENDFOR
%\FORALL{$j \in S_i$}
%\IF{$j \notin D$}
%\FORALL{$k \in S_j$}
%\IF[$i$ and $j$ mutually influenced by $k$]{$k$ is marked NEW-C-POINT}
%\STATE $w_j \leftarrow w_j - 1$ \COMMENT{see
%  Figure~\ref{3:fig:CLJP-update-heuristics} right}
%\STATE $S_i = S_i \setminus \{j\}$ \COMMENT{remove edge from $j$ to $i$}
%\ENDIF
%\ENDFOR
%\ENDIF
%\ENDFOR
%\FORALL{$k$ originally in $S_i$}
%\IF{$k \in D$}
%\STATE unmark $k$
%\STATE $S_i = S_i \setminus \{k\}$ \COMMENT{remove edge from $k$ to $i$,
%  if present}
%\ENDIF
%\ENDFOR
%\ENDFOR
%\end{algorithmic}
%\}
%\end{algorithm}
The level of complication in this update routine is due to the CSR
format and the need to find vertices strongly influenced by new
$C$-points. When a new $C$-point $k$ is selected, the first type of
weight update (Lines 2--7) is trivial since determining the vertices
in $S_k$ is inexpensive. The second type of update (Lines 8--30) is
more expensive since the vertices influenced by $k$ are difficult to
determine in a CSR format. The update requires a search of many
vertices $i$ and all of their strong influencing neighbors $j$. The
routine then searches strongly influencing $j$ to determine if any $k
\in D$ strongly influences both $i$ and $j$. The cost increases
dramatically as the density of $S$ increases. Large operator
complexities have a disproportionately large impact on coarse-grid
selection run time. In the next section, a modified update routine to
compliment the new search technique is introduced.
